---
title: "ERP visualization with three conditions"
output: html_document
layout: post
comments: true
categories: [EEG, ERPs, statistics, R, ggplot2]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dev = 'svg')
options(contrasts = c("contr.sum","contr.poly"))
```

In part 1 I took a look at visualizing ERPs from two conditions at a single electrode. This time I'm going to look at three conditions. As in the previous post, I'll assume a basic familiarity with ERPs.

First I'll load in the full dataset, which contains ERPs for all conditions for all subjects and whip it into shape.

```{r preProc}
library(ggplot2)
library(reshape2)
library(dplyr)
levCatGAall <- read.csv("https://raw.githubusercontent.com/craddm/ExploringERPs/master/levCatGAall.csv",header=FALSE)
names(levCatGAall) <- c("ObjFull","ObjHP","ObjLP","NonFull","NonHP","NonLP","Time","Subject")
levCatGAall <- levCatGAall[(levCatGAall$Time >= -100)& (levCatGAall$Time <= 400),]
levCatGAall$Subject <- as.factor(levCatGAall$Subject)
levCatGAall <- melt(levCatGAall,id.vars = c("Subject","Time"))
names(levCatGAall) <- c("Subject","Time","condition","amplitude")
levCatGAall$Object <- gl(2,11520,labels = c("Object","Non-Object"))
levCatGAall$Frequency <- gl(3,3840,labels = c("BB","HSF","LSF"))

ERP.plot <- ggplot(levCatGAall,aes(Time,amplitude))+
  scale_color_brewer(palette = "Set1")+
  theme_classic()
```

## The Plots

Let's start off with a basic, standard ERP plot.

```{r basicPlot, fig.height = 5, fig.width = 7}
ERP.plot +
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = Frequency)) +
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "") +
  geom_vline(xintercept = 0,linetype = "dashed" ) +
  geom_hline(yintercept = 0,linetype = "dashed")
```

Looks pretty much like there's not a lot of difference in the P1, but HSF and BB are more negative in the N1 and P2. 

Let's add bootstrapped 95% confidence intervals. Note that you could add SE instead of confidence intervals by replacing **mean_cl_boot** with **mean_se**.

```{r CIPlot,fig.height = 5, fig.width = 7}
ERP.plot+
  stat_summary(fun.data = mean_cl_boot,geom = "ribbon",size = 1,aes(fill = Frequency),alpha = 0.3)+
  guides(fill = "none")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = Frequency),alpha = 0.8)+
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")
```

There's a lot of overlap here, so the picture is already starting to get muddy. As noted in the previous post, the CIs are not terribly useful for inference here because they show between-subject variability when what's interesting is the within-subject variability, the variability of the differences. Nevertheless, we'll plow on adding individual subject data.

```{r indivAndGroup, fig.height = 5, fig.width = 8}
ERP.plot +
  stat_summary(fun.y = mean,geom = "line",alpha = 0.4,aes(group =interaction(Subject,Frequency),colour = Frequency),size = 0.7) +
  guides(alpha= "none") +
  stat_summary(fun.data = mean_cl_boot,geom = "ribbon",alpha = 0.4,aes(fill = Frequency)) +
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = Frequency),alpha = 0.8) +
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "") +
  guides(colour= "none") +
  geom_vline(xintercept = 0,linetype = "dashed") +
  geom_hline(yintercept = 0,linetype = "dashed")
```

Now it's becoming a real mess. It's really difficult to tell the conditions apart properly in the individual traces. Poor choice of colours? Maybe, but I tried rendering this plot under multiple colour schemes and it simply didn't help. And this problem can only get worse when you add more conditions. Let's split the conditions up.
  
```{r condSplit, fig.height = 5, fig.width = 8}
ERP.plot+
  facet_wrap(~Frequency)+
  stat_summary(fun.y = mean,geom = "line",aes(group = Subject),alpha = 0.3)+
  stat_summary(fun.data = mean_cl_boot,geom = "ribbon",alpha = 0.4)+
  guides(fill= "none")+
  stat_summary(fun.y = mean,geom = "line",size = 1)+
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")
```

Now we have three subplots, each informative about a particular condition. The mental gymnastics to work out where the differences are are becoming harder, and I find myself relying on the condition means. Couple of points of note - you have a couple of people who look a bit odd in the 160-200 ms time window - where most people are going negative, they aren't. After 230 ms or so, the data in general seems much more variable. But since there's no way to tell which line belongs to which participant, it's hard to really see what I gain here when I'm interested in the within-participant differences. Again, difference waves might be more helpful.

## Difference waves

In the two condition post, life was simple. Two conditions only need a single difference wave. But as the number of conditions increases, the number of pairwise differences increases. For three conditions, you'd need three difference waves (BB-HSF, BB-LSF, HSF-LSF). For four conditions, you'd need six difference waves. But of course, you'd start off with an F-test to test whether any of the means differ from each other, and, if they do, we'd then run post-hoc t-tests to check which pairs of means differ. A simple starting point is thus the difference between each condition mean and the *grand* mean, the mean across all conditions. 

```{r diffPlotGroup, fig.height = 5, fig.width = 7}
levCatGAall$GA <- ave(levCatGAall$amplitude,levCatGAall$Time)
levCatGAall$GAdiff <- levCatGAall$amplitude-levCatGAall$GA

ERPdiff.plot <- ggplot(levCatGAall,aes(Time,amplitude))+
  theme_classic()

ERPdiff.plot+
  guides(fill = "none")+
  labs(x = "Time (ms)", y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,fill = Frequency),alpha = 0.8)+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")

```

In theory, we could add lines for each person's difference from the grand mean...

```{r diffPlotIndiv}
ERPdiff.plot+
  guides(fill = "none")+
  labs(x = "Time (ms)", y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,group = interaction(Subject,Frequency),colour = Frequency),alpha = 0.4)+
  stat_summary(fun.y = mean,geom = "line",size = 2,aes(y = GAdiff,colour = Frequency))+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")
```

...but if you can make sense of that, you have better eyesight than I have.


```{r diffPlotIndiv2}
ERPdiff.plot+
  guides(fill = "none")+
  labs(x = "Time (ms)", y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,group = interaction(Subject,Frequency),colour = Frequency),alpha = 0.4)+
  stat_summary(fun.y = mean,geom = "line",size = 2,aes(y = GAdiff,colour = Frequency))+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")
```

```{r diffPlot2}
ERPdiff.plot+
  guides(fill = "none")+
  labs(x = "Time (ms)", y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  stat_summary(fun.data = mean_se,geom = "ribbon",size = 1,aes(y = GAdiff,fill = Frequency),alpha = 0.4)+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,colour = Frequency))+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")
```


```{r addSig}

library(afex)
library(purrr)


FtestPurrAfx <- levCatGAall %>%
  split(.$Time) %>%
  map(~ aov_ez("Subject","amplitude",.,within = c("Object","Frequency"))) %>%
  map_dbl(.,c(1,6,2))
  
levCatGAall$pval <- p.adjust(FtestPurrAfx,"none")
levCatGAall$pval[levCatGAall$pval > .05] <- NA

ERPdiff.plot <- ggplot(levCatGAall,aes(Time,amplitude))+
  theme_classic()

ERPdiff.plot+
  guides(fill = "none")+
  labs(x = "Time (ms)", y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
    stat_summary(fun.data = mean_cl_boot,geom = "ribbon",size = 1,aes(y = GAdiff,fill = Frequency),alpha = 0.4)+
  #stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,fill = Frequency),alpha = 0.4)+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(y = GAdiff,colour = Frequency))+
  stat_summary(fun.y = mean,geom= "point",aes(y = pval-2))+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")

```


As you can probably see from the above, with just three conditions, some of the recommendations for improving standard ERP figures start to fall down, in my opinion. 