---
title: "ERP Visualization: Within-subject confidence intervals"
output: html_document
layout: post
comments: true
categories: [EEG, ERPs, statistics, R, ggplot2]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,dev = 'svg')
```

In a previous post I implemented running t-tests and straightforward methods of control for multiple comparisons - Bonferroni-Holm and FDR. Another method of correcting for multiple comparisons is the non-parametric cluster-based permutation test. Let's prep the data and load all the packages I'll need.

```{r prepData, message = FALSE}
library(tidyverse)
library(ggplot2)

levCatGA <- read_csv("https://raw.githubusercontent.com/craddm/ExploringERPs/master/levCatObjNon.csv",col_names = c("Object", "Non-Object", "Time", "Subject"))
levCatGA$Subject <- as.factor(levCatGA$Subject)
levCatGA <- levCatGA[(levCatGA$Time >= -100)& (levCatGA$Time <= 400),] %>%
  gather(condition,amplitude,-Subject,-Time)

theme_set(theme_classic())

levCat.plot <- ggplot(levCatGA,aes(Time,amplitude))+
  scale_color_brewer(palette = "Set1")
```

Now let's run t-tests on each timepoint (again, using [***purrr***](../blog/2016/10/06/2016/10/06/ERP-Vis-Running-Tests)) 

```{r runTests, message = FALSE, warning = FALSE}
runningT <- levCatGA %>%
    split(.$Time) %>%
    map(~t.test(amplitude~condition,paired = TRUE,data = .))
```

I now have two lists, **runningT**, containing the t-test result for each timepoint, and **runningSE**, containing the summarized data from each time point and the corrected SEs/CIs. For example:

```{r someTests}
runningT$`0`
runningSE$`0`
```

As before, I'll extract the p-value for each t-test and store it in a custom data-frame for later plotting. Let's plot the ERPs with standard between-subject confidence intervals.

```{r betweenSubj,fig.height = 4, fig.width = 6}

pvals <- data.frame(
  Time = unique(levCatGA$Time),
  p.value = map_dbl(runningT,"p.value")
  )

pvals$crit <- 0+(pvals$p.value <= .05)
pvals$crit[pvals$crit == 0] <- NA

levCat.plot+
  stat_summary(fun.data = mean_cl_boot,geom = "ribbon",aes(fill = condition,colour = condition),linetype = "dashed",alpha = 0.3)+
  guides(fill = "none")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = condition))+
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  geom_line(data = pvals, aes(x = Time, y = crit-3),na.rm = TRUE,size = 2)+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")

```

The confidence intervals in the between-subjects plot don't really help you see that the signficant timepoints are actually significant. Let's re-plot the figure using within-subject CIs. I use here some new commands. Using **map_df** from ***purrr***, I convert runningSE, a list of data frames, into a single data frame summarizing the data at each time-point after removing between-subject variability. This also uses **extract** function from ***magrittr***. I then pass this data frame to geom_ribbon to plot the corrected CIs. I could also use it to plot the ERPS, but I'll just let ggplot figure them out from the original data.


```{r withinCI,fig.height = 4, fig.width = 6}

WSCI <- map_df(runningSE,extract) %>%
  mutate(
    Time = rep(unique(levCatGA$Time),each =2) #Note, you'll have to change 2 to match the number of conditions
    )
  
levCat.plot+
  geom_ribbon(data = WSCI,aes(ymin = amplitude-ci,ymax = amplitude+ci,fill = condition,colour = condition),linetype="dashed",alpha = 0.3)+
  guides(fill = "none")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = condition))+
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  geom_line(data = pvals, aes(x = Time, y = crit-3),na.rm = TRUE,size = 2)+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")

```

These confidence intervals are much narrower, reflecting the correlation between measures within participants. These intervals make it easier to see where significant differences lie from the degree of overlap between. 

Finally, let's overlap the two types of confidence interval. 

```{r bothTypes,fig.height = 4, fig.width = 6}
levCat.plot+
  stat_summary(fun.data = mean_cl_normal,geom = "ribbon",aes(colour = condition),fill = NA,linetype = "dotted",alpha = 0.3)+
  geom_ribbon(data = WSCI,aes(ymin = amplitude-ci,ymax = amplitude+ci,fill = condition,colour = condition),linetype="dashed",alpha = 0.3)+
  guides(fill = "none")+
  stat_summary(fun.y = mean,geom = "line",size = 1,aes(colour = condition))+
  labs(x = "Time (ms)",y = expression(paste("Amplitude (",mu,"V)")),colour = "")+
  geom_line(data = pvals, aes(x = Time, y = crit-3),na.rm = TRUE,size = 2)+
  geom_vline(xintercept = 0,linetype = "dashed" )+
  geom_hline(yintercept = 0,linetype = "dashed")

```

Filled, dashed lines indicate within-subject confidence intervals, while the outer, dotted lines show the position of the between-subject confidence intervals. As we saw above, the WSCIs are much narrower, largely due to the correlation between measures from within-participants.

So be careful interpreting differences from confidence intervals - check whether they show within or between subject differences before using them for Eyeball Mark I inference.